A Global Picture of Hunger

### 2020-11-18 docs/modeling_fies.tex
new draft with country-level leave-one-out cross validation

\maketitle
\begin{abstract}
Using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the \href{https://worldhunger.io}{World Hunger Clock}.  This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.  We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population food insecure in some high-income countries to the developing world where over half the population is severely food insecure.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been steadily declining, with large observed and forecasted gains in lower and lower-middle income countries, and decreases in total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.  Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population experiencing hunger by 2030.  \end{abstract}

\section{Main}
Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate to severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES. The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger. Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population. These do not directly capture the actual lived exposure of individuals to food shortages.

@@

This paper contributes to shifting the research frontier by filling these two gaps in terms of data coverage and forecasts by developing a statistical model of FIES outcomes based on covariates that are available for territories where actual polling data are not available. Such a model can be employed to obtain forecasts to 2030 and assess likely future trends in food insecurity. We take advantage of individual-level characteristics collected in FIES surveys to push the analysis beyond country level to a sub-national level. To our knowledge, this is the first effort to model the FIES outcomes at that scale.

Food security has traditionally been difficult to measure, and this has led to an incomplete or inaccurate assessment of global hunger. Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.  However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \citep{Perumal2018}.  Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.  Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.  Thus, global estimates of hunger and food insecurity based on these metrics carry forward similar flaws.

@@

\section{Results}
\subsection{Food Insecurity over Time}
Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and an additional 2.5 billion people are above the threshold for moderate food insecurity (See Fig. \ref{fig:timeseries}). These numbers are higher than the estimates in the 2019 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \ref{sofi2020}.

Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.  However, for both indicators, these changes are very slight: from 2010 to 2030, the number of moderately food insecure people will increase 2\%, and the number of severely food insecure people will decrease 8\%.  The increases in moderate food insecurity have been in lower and lower-middle income countries, while the increases in severe food insecurity are entirely from lower income countries.  For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.

@@

Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig. \ref{fig:rates}, we find clear improvements in all except high income countries.  Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.  Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.  In spite of this progress in the bottom three income quartiles, in high-income countries, rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.  Overall, the world has made steady but slow progress on reducing the rate of food insecurity.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/Rates.pdf}
  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by World Bank income groups as of 2020}
  \label{fig:rates}
\end{figure}

@@

\begin{landscape}
\begin{figure}[h]
  \centering
  \includegraphics{img/FullMap.png}
  \caption{Spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
  \label{fig:map}
\end{figure}
\end{landscape}

\subsection{Important Predictors of Food Insecurity}
Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.950$ and $R^2 = 0.944$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.799$ and $R^2 = 0.772$ for moderate and severe food insecurity, respectively).  The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the coronavirus pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.  Nevertheless, lower and lower-middle income countries, are expected to see overall increases in the number of food insecure people.

Our analysis of the relative importance of different predictors for future food insecurity dynamics shows that economic conditions are the largest driver of global variation in rates of food insecurity. For both moderate and severe food insecurity, the Poverty Headcount Index, the Gini coefficient, and GDP per capita appear as highly relevant predictors of the rate of food insecurity. The rate of stunting, an indicator of both sanitary conditions and chronic, long-term hunger is also an important predictor of food insecurity. However, other indicators of sanitary conditions, such as the rates of wasting and malaria mortality are less important as explanatory factors of changes in food insecurity. Geographic and climate factors, such as the percentage of urban population, topographic ruggedness and mean annual temperature or precipitation do not appear to be robust predictors of food insecurity at a global scale. Finally, while economic conditions are widely associated with food insecurity at a global scale, the fact that high income countries have seen little improvement in food insecurity outcomes, in spite of marginal annual increases in GDP per capita, suggests that there is an upper limit to the extent to which increasing economic development can improve food security outcomes.

Our results anticipating of increases in the number of people over the threshold for moderate food insecurity concur with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.  However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.  In this case, our predictions diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 3\% decrease, from 788 to 767 million severely food insecure people over the same period.

@@

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.  We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 3-7 percentage points on average when applied to countries outside of the training set.

@@

\section{Methods}
\subsection{Disaggregation}
The data on the FIES collected by Gallup records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  We use a similar methodology and calculate post-stratification weights at a subnational level.  We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}, where available.  This methodology rests on the mild assumption that subnational differences in rates of food insecurity are driven by factors related to demographics, urbanization, and wealth rather than by other unobservable factors.

\subsection{Covariates}
We model subnational rates of severe and moderate-to-severe food insecurity as a function of several covariates related to human development, food security, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).  We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.  For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.

@@

Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.  These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.  For a detailed overview of the steps involved in processing and preparing each covariate, see the Appendix.

@@

When performing cross validation, it is typical to select observations at random.  However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.  Thus, for each iteration of cross validation, we leave out data from 10\% of the countries in our dataset, rather than 10\% of the total observations.  Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.  By this validation metric, we find that our models perform well, with an $R^2$ of 0.799 and mean average error (MAE) of 0.068 for moderate food insecurity and an $R^2$ of 0.772 and a MAE of 0.038 for severe food insecurity.  After fitting models on the entire dataset, we observe an $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.944 and a MAE of 0.019 for severe food insecurity.

After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable, breiman2001random}.  These metric can provide some interpretability to model predictions.

The Supplementary Materials provide a detailed description of the implementation and validation of the random forest models.

@@

>>>>

\maketitle
\begin{abstract}
Using microdata from 75 countries collected by the FAO and Gallup World Poll for the Voices of the Hungry project, we modeled levels of food security at the subnational level from 2010 to 2030 to create the \href{https://worldhunger.io}{World Hunger Clock}.  This is the first global picture at a subnational level of the Food Insecurity Experience Scale, the food security metric most indicative of the lived experience of hunger.  We find significant heterogeneity in food security around the world, ranging from less than 4\% of the population food insecure in some high-income countries to the parts of developing world where over half the population is severely food insecure.  Examining global temporal trends and accounting for the effects of the COVID-19 pandemic, we find that rates of food insecurity have been steadily declining, with large observed and forecasted gains in lower and lower-middle income countries, and decreases in total number of severely food insecure people.  However, the total number of moderately food insecure people has been increasing and, after recovering from the shock of the COVID-19 pandemic, will continue to increase through the end of the 2020s.  Overall, we conclude that global gains have been incremental, and current trends in development and demographic change will still leave a large share of the world's population experiencing hunger by 2030.  \end{abstract}

\section{Main}
Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of at least moderate food insecurity in the population based on the Food Insecurity Experience Scale, or FIES. The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger. Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population. These do not directly capture the actual lived exposure of individuals to food shortages.

@@

This paper contributes to shifting the research frontier by filling these two gaps in terms of data coverage and forecasts by developing a statistical model of FIES outcomes based on covariates that are available for territories where actual polling data are not available. Such a model can be employed to obtain forecasts to 2030 and assess likely future trends in food insecurity. We take advantage of individual-level characteristics collected in FIES surveys to push the analysis beyond the country level to a sub-national level. To our knowledge, this is the first effort to model the FIES outcomes at that scale.

Food security has traditionally been difficult to measure, and this has led to an incomplete or inaccurate assessment of global hunger. Metrics of macro-health, such as anthropometric measures and mortality rates are correlated broadly with food insecurity and have been used for many years to monitor human well-being \citep{Puffer1973, Habicht1974}.  However, these metrics are affected by other determinants of health such as the occurrence of infectious diseases, and are not meaningful at the scale of individuals or households \citep{Perumal2018}.  Other proxies for food security, such as food availability estimated from crop yields \citep{Maxwell1992}, are also inadequate because they only make rough estimates of how accessible food is to the general population, and can elide populations that are food insecure due to poor access, even when aggregate food availability is high \citep{Sen1983}.  Moreover, these metrics are very sensitive to incorrect estimates of crop yields and food reserves at a national scale.  Thus, global estimates of hunger and food insecurity based on these metrics carry forward similar flaws.

@@

\section{Results}
\subsection{Food Insecurity over Time}
Globally, in the year 2020, we estimate that 820 million people are above the threshold for severe food insecurity, and 2.5 billion people are above the threshold for moderate food insecurity (See Fig. \ref{fig:timeseries}). These numbers are higher than the estimates in the 2020 FAO State of Food Insecurity report, which reports about 746 million severely food insecure people and 2 billion people experiencing at least moderate food insecurity in 2019 \ref{sofi2020}.

Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.  However, for both indicators, these changes are very slight: in 2030, the number of moderately food insecure people will have increased by 2.2\% since 2010, and the number of severely food insecure people will have decreased by 7.0\%.  The increases in moderate food insecurity have been in lower and lower-middle income countries, while the increases in severe food insecurity are entirely from lower income countries.  For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.

@@

Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig. \ref{fig:rates}, we find clear improvements in all except high income countries.  Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.  Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.  In spite of this progress in the bottom three income quartiles, in high-income countries, the rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.  Overall, the world has made steady but slow progress on reducing the rate of food insecurity.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/Rates.pdf}
  \caption{Percentage of the population over the threshold for moderate and severe food insecurity, by World Bank income groups as of 2020.}
  \label{fig:rates}
\end{figure}

@@

\begin{landscape}
\begin{figure}[h]
  \centering
  \includegraphics{img/FullMap.pdf}
  \caption{Spatial distribution of moderate and severe food insecurity for the years 2010, 2020, and 2030.}
  \label{fig:map}
\end{figure}
\end{landscape}

\subsection{Important Predictors of Food Insecurity}
Overall, our model was able to predict food insecurity with high accuracy in the countries for which training data was available ($R^2 = 0.950$ and $R^2 = 0.940$ for moderate and severe food insecurity, respectively), and excluding entire countries from the training data to evaluate out-of-country model error still showed overall good results ($R^2 = 0.792$ and $R^2 = 0.746$ for moderate and severe food insecurity, respectively).  The relative importance of different variables as predictors of food insecurity can be assessed by examining the loss in forecast accuracy resulting from the exclusion of particular covariates.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the coronavirus pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.  Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of food insecure people.

Our analysis of the relative importance of different predictors for future food insecurity dynamics shows that economic conditions are the largest driver of global variation in rates of food insecurity. For both moderate and severe food insecurity, the Poverty Headcount Index, the Gini coefficient, and GDP per capita appear as highly relevant predictors of the rate of food insecurity. The rate of stunting, an indicator of both sanitary conditions and chronic, long-term under-nutrition is also an important predictor of food insecurity. However, other indicators of nutrition and sanitary conditions, such as the rates of wasting and malaria mortality are less important as explanatory factors of changes in food insecurity. Geographic and climate factors, such as the percentage of urban population, topographic ruggedness and mean annual temperature or precipitation do not appear to be robust predictors of food insecurity at a global scale.

While economic conditions, and especially GDP per capita, are important predictors of food insecurity at a global scale, there is an upper limit to the extent to which increasing economic development alone can improve food security outcomes.  High income countries have seen little improvement in food insecurity outcomes over twenty years, and upper-middle income countries are expected to see diminishing gains throughout the 2020s.  Among the high-income countries in our dataset, those with the lowest rates of moderate and severe food insecurity also had lower inequality and poverty.  This suggests that our model's forecasted marginal increases in food insecurity in high income countries is due to trends of increasing income inequality, and that economic development alone cannot ensure that everyone is well-fed.

Our results anticipating increases in the number of people over the threshold for moderate food insecurity concur with the FAO's predictions of increased rates of undernourishment, the other indicator for the SDG 2 target of ending hunger \citep[Part 1, Page 11]{sofi2020}.  However, severe food insecurity may be a more apt indicator to compare with undernourishment, as the global totals are more similar \citep{sofi2020}.  In this case, our predictions diverge: the FAO estimates a 24\% increase, from 687 to 841 million undernourished people, from 2019 to 2030, whereas we predict a 2.6\% decrease, from 787 to 767 million severely food insecure people over the same period.

@@

When making predictions in countries where we do not have primary data, we assume that the relationships among variables like poverty, stunting, inequality, and hunger are the same across all locations.  We test the accuracy of these cases by using ten-fold cross validation at the country level, and we find good accuracy, with our models making estimates that are only off by 4-7 percentage points on average when applied to countries outside of the training set.

@@

\section{Methods}
\subsection{Disaggregation}
The data on the FIES collected by Gallup records many individual-level attributes from the respondents, including age, gender, wealth quintile, and whether they live in an urban or rural area.  Using national data on these variables, a standard weighting scheme is created for each individual, based on the ratio of population probabilities to sample probabilities of an individual with those characteristics being selected \citep{bethlehem2009applied}.  We use a similar methodology and calculate post-stratification weights at a subnational level.  We use year-specific subnational estimates of population shares by gender, age, urbanization, and, where available, wealth, and create a separate set of weights for each subnational area across all individuals.  For age and gender, we used gridded data from WorldPop \citep{Tatem2017}; for urbanization, we used spatially explicit estimates published by Jones and O'Neill \citep{Jones2016}; and for wealth we used data on household wealth quintiles from Demographic and Health Surveys \citep{dhsall}.  This methodology rests on the mild assumption that, within a given country, two individuals of the same age, gender, urbanization level and wealth quintile will have a similar food security status.

\subsection{Covariates}
We model subnational rates of moderate and severe food insecurity as a function of several covariates related to human development, nutrition, health, infrastructure, income levels and distribution, and the environment (See Table \ref{tab:covars}).  We employ covariates for which projections are available up to year 2030 and, in most cases, also have subnational spatial resolution.  For projections based on the Shared Socioeconomic Pathways (SSP) framework \citep{oneill2014new}, we use the middle-of-the-road pathway, SSP2, and for climatological variables, we use forecasts based on the Representative Concentration Pathway (RCP) 6.0 \citep{van2011representative}.

@@

Finally, we adjust several variables to account for the effects of the COVID-19 pandemic.  These include June 2020 GDP predictions from the World Bank \citep{prospects2020}, August 2020 estimates of impacts on anthropometry from \textit{The Lancet} \citep{headey2020impacts}, as well as the Poverty Headcount Index \cite{Cuaresma2018}.

@@

When performing cross validation, it is typical to select observations at random.  However, in our training data we have multiple highly similar observations from each country across years and subnational areas, and we extend our model predictions to countries that were not found at all in our training data.  Thus, we perform a type of leave-one-out cross validation, where we leave out all of the data from one country for each iteration.  Thus, the model error after cross validation gives us a sense of how the model is performing in countries where FIES data was not collected.  By this validation metric, we find that our models perform well, with an $R^2$ of 0.792 and mean average error (MAE) of 0.071 for moderate food insecurity and an $R^2$ of 0.746 and a MAE of 0.041 for severe food insecurity.  After fitting models on the entire dataset, we observe a within-sample $R^2$ of 0.950 and an MAE of 0.034 for moderate food insecurity and an $R^2$ of 0.945 and a MAE of 0.018 for severe food insecurity.

After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in Mean Average Error (MAE) between the regression \citep{ishwaran2007variable, breiman2001random}.  This metric can provide some interpretability to model predictions.

The Supplementary Materials provide a detailed description of the data processing for the covariates, as well as implementation and validation of the random forest models.

\section

@@

\end{document}

### 2020-11-18 docs/modeling_fies_supplement.tex
new draft with country-level leave-one-out cross validation

\section{Overview of the FIES}
We use raw microdata released by the FAO from 75 countries from different world regions and income categories (See Fig. \ref{fig:fies_countries}.

@@

Based on the global reference scale, two thresholds are set based on questions 1 and 8: whether the respondent is worried they would not have enough food to eat because of a lack of resources, and whether there was a time when they went without eating for a whole day because of a lack of money or other resources.

\begin{enumerate}
	\item During the last 12 MONTHS, was there a time when you were worried you would not have enough food to eat because of a lack of money or other resources?
	\item Still thinking about the last 12 MONTHS, was there a time when you were unable to eat healthy and nutritious food because of a lack of money or other resources?
	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
	\item Still thinking about the last 12 MONTHS, was there a time when you ate less than you thought you should because of a lack of money or other resources?
	\item Was there a time when your household ran out of food because of a lack of money or other resources?
	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
	\item During the last 12 MONTHS, was there a time when you went without eating for a whole day because of a lack of money or other resources?
  \label{itm:fies}
\end{enumerate}

@@

Each ROC is weighted so as to give proportionally more weight to recent years.  For a dataset with observations for 2010 to 2017, that would be:

\begin{equation}
  W_y = \frac{y - 2010}{\sum_{y=2010}^{2017} y - 2010}
  \label{eqn:b}
\end{equation}

@@

\begin{equation}
  \text{AROC}_{a} = \text{logit} \left( \sum_{y=2010}^{2017} W_y \times \text{AROC}_{a, y} \right)
  \label{eqn:c}
\end{equation}

@@

\begin{equation}
  \text{Proj}_{a,y} = \text{logit}^{-1} ( \text{logit} ( p_{a,2017} ) + AROC_{a} \times ( y - 2017 ) )
  \label{eqn:d}
\end{equation}

@@

\pagebreak
\section{Description, Implementation and Validation of the Random Forest Regression Model}
The RF regression is a machine learning method based on the creation of a large number of decision trees. By taking an ensemble of decision tree models, random forests introduce more variance and balance out the bias that is common to methods based on single decision trees \citep{friedman2001elements}.  For each of the decision trees in a random forest model, observations are selected at random with replacement, a method known as bootstrap aggregating, or bagging, and features are also selected at random.

For the implementation we use the R-package \texttt{randomForestSRC} by Ishwaran and Kogalur \citep{ishwaran2019randomforestsrc}. Specifically, we apply the functions \texttt{tune.rfsrc()} to find the optimal combination of hyperparamters and the function \texttt{rfsrc()} to perform the RF regression models. After predicting rates of moderate-to-severe and severe food insecurity, we plot the modeled vs. observed rates and calculated the MSE and ${R}^2$ (See Fig. \ref{fig:rf_in-sample} and Fig. \ref{fig:rf_out-sample}) for both models. Additionally, we use the function \texttt{vimp()} to gain insights into the development of the error rates with increasing numbers of trees (See Fig. \ref{fig:rf_error}) and the importance of individual variables (See Fig. \ref{fig:rf_vimp}).

%STEP-BY-STEP: LOGIT TRANSFORM, TUNE, MODEL, PREDICT, VALIDATE
Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.  Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.

To ensure that predicted rates are bounded by 0 and 1 we conduct a logit transformation of the dependent variable and inverse logit transformation of the predictions. 
%MAYBE STATE THIS SOMEWHERE ELSE, NOT SURE WHERE

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/error_rf.png}
  \caption{Error Rate of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_error}
\end{figure}

%HERE A FEW SENTENCES ON THE NUMBER OF TREES WITH LINK TO NEXT PARAGRAPH

In addition to the parameter that describes the number of trees to be created, two other important hyperparameters must be tuned. In the function \texttt{tune.rfsrc()} these parameters are called \texttt{mtry} and \texttt{nodesize}. \texttt{mtry} describes the number of variables randomly selected as candidates for splitting a node and \texttt{nodesize} describes the average number of observations in a leaf node. The optimal pair of hyperparamters is found by trying different combinations and choosing the one with the smallest out-of-bag (OOB) error.

%EXPLAIN THE OOB ERROR HERE?

The function \texttt{tune.rfsrc()} does exactly that and returns optimal values for \texttt{mtry} and \texttt{nodesize}. We applied this function on both, the moderate-to-severe and the severe model. For the moderate-to-severe model the optimal combination is \texttt{mtry} = 12 and \texttt{nodesize} = 1 and for the severe model we found \texttt{mtry} = 10 and \texttt{nodesize} = 2 to be optimal.

%STATE FUNCTION ARGUMENTS HERE?

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/in-sample_rf.png}
  \caption{In-Sample Fit of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_in-sample}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/model/out-sample_rf.png}
  \caption{Out-Of-Sample Fit of the RF regression model with 75\% training set and 25\% test set. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_out-sample}
\end{figure}

%HERE EVERYTHING ON THE MODEL, NOT SURE HOW DETAILED
%For modeling the rates of moderate-to-severe and severe food insecurity we use the function \texttt{rfsrc()}. We specify

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/VIMP.pdf}
  \caption{Variable Importance of the RF regression model. Panel (A) shows the Moderate-to-Severe model and Panel (B) shows the Severe model.}
  \label{fig:rf_vimp}
\end{figure}

%https://kogalur.github.io/randomForestSRC/theory.html#section4.2
Lastly we asses the importance of individual variables. This is based on the Breiman-Culter permutation variable importance information \citep{breiman2001random}.  This involved comparing the prediction error on the OOB data to the prediction error of OOB cases where the x-variable $x$ is randomly permutated.  The importance of an individual variable is thus the mean difference between the perturbed and unperturbed error rate, across all trees. 
%MORE INFORMATION HERE
\end{document}

>>>>

\section{Overview of the FIES}
We use raw microdata released by the FAO from 75 countries from different world regions and income categories (See Fig. \ref{fig:fies_countries}).

@@

Based on the global reference scale, two thresholds are set based on questions 1 and 8, corresponding to moderate and severe food insecurity, respectively.

\begin{enumerate}
	\item During the last 12 months, was there a time when you were worried you would not have enough food to eat because of a lack of money or other resources?
	\item Still thinking about the last 12 months, was there a time when you were unable to eat healthy and nutritious food because of a lack of money or other resources?
	\item Was there a time when you ate only a few kinds of foods because of a lack of money or other resources?
	\item Was there a time when you had to skip a meal because there was not enough money or other resources to get food?
	\item Still thinking about the last 12 months, was there a time when you ate less than you thought you should because of a lack of money or other resources?
	\item Was there a time when your household ran out of food because of a lack of money or other resources?
	\item Was there a time when you were hungry but did not eat because there was not enough money or other resources for food?
	\item During the last 12 months, was there a time when you went without eating for a whole day because of a lack of money or other resources?
  \label{itm:fies}
\end{enumerate}

@@

Each ROC is weighted so as to give proportionally more weight to recent years.  For a dataset with observations for 2010 to 2019, that would be:

\begin{equation}
  W_y = \frac{y - 2010}{\sum_{y=2010}^{2019} y - 2010}
  \label{eqn:b}
\end{equation}

@@

\begin{equation}
  \text{AROC}_{a} = \text{logit} \left( \sum_{y=2010}^{2019} W_y \times \text{AROC}_{a, y} \right)
  \label{eqn:c}
\end{equation}

@@

\begin{equation}
  \text{Proj}_{a,y} = \text{logit}^{-1} ( \text{logit} ( p_{a,2019} ) + AROC_{a} \times ( y - 2019 ) )
  \label{eqn:d}
\end{equation}

@@

\pagebreak
\section{Description, Implementation and Validation of the Random Forest Regression Model}
The random forest regression is a machine learning method based on the creation of a large number of decision trees.  For each of the decision trees in a random forest model, observations are selected at random with replacement, a method known as bootstrap aggregating, or bagging, and features are also selected at random.  For the implementation we use the R-package \texttt{randomForestSRC} by Ishwaran and Kogalur \citep{ishwaran2019randomforestsrc}.

Because the outcome variables in our models, the rate of moderate or severe food insecurity, are a bounded outcome, we first used a logistic transformation to extend the range from $[0,1]$ to $(-\infty, \infty)$.  Then, after estimating the model, we used the inverse logit to convert the unbounded model predictions to a rate between 0 and 1.

To determine the best hyperparameters to model both moderate ands severe food insecurity, we use leave-one-out cross validation at the country level.  This involves running a separate model for each country, where we train a model on the remaining data from all of the other countries, and use that model to predict values for the country that was left out.  Then, across the predicted values for all countries, we evaluate the mean average error (MAE) and $R^2$ of the differences between observed rates of food insecurity and predictions from the cross validation models.

We evaluate models across a grid of hyperparameters, with combinations of: between 1 to 13 variables randomly selected as candidates for splitting a node (\texttt{mtry}), between 1 and 20 terminal node sizes (\texttt{nodesize}), and with a maximum tree depth (\texttt{depth}) between 1 to 13 nodes, as well as without any maximum tree depth.  For each model in cross validation, we used 1000 trees (\texttt{ntree}).  Due to the large size of the hyperparameter space, we sampled sets of hyperparameters randomly, eventually sampling about 50\% of all combinations of hyperparamters.  We found that the optimal model for moderate food insecurity had 16 average terminal nodes, 2 variables selected as candidates for splitting a node, and no maximum tree depth; the optimal model for severe food insecurity also had 2 variables selected as candidates for splitting a node and no maximum tree depth, but with only 14 average terminal.

After fitting models with cross validation, we evaluated the models performance and graphed a comparison between the observed and predicted values based on instances where individual countries were left out (See Fig. \ref{fig:rf_out-sample}), as well as for a model trained on the entire dataset, which can only be evaluated on in-sample data (See Fig. \ref{fig:rf_in-sample}).  Finally, Figure \ref{fig:mod-error} shows the model error converging and the number of trees increases.

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/out-sample_rf.pdf}
  \caption{Out-of-sample fit of the random forest regression model, where predictions for each country are based on a model that did not include that country in the training data. Panel (A) shows the model for moderate food insecurity and panel (B) shows the model for severe food insecurity.}
  \label{fig:rf_out-sample}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/in-sample_rf.pdf}
  \caption{In-sample fit of the random forest regression model, where one model was trained on the entire dataset. Panel (A) shows the model for moderate food insecurity and panel (B) shows the model for severe food insecurity.}
  \label{fig:rf_in-sample}
\end{figure}

\begin{figure}[H]
  \centering
  \includegraphics[width=\linewidth]{img/Error_Over_Training.pdf}
  \caption{Error rate of the random forest regression model as the number of trees in the model increases. Panel (A) shows the model for moderate food insecurity, and panel (B) shows the model for severe food insecurity.}
  \label{fig:mod-error}
\end{figure}

\end{document}
